---
title: "Solution for lab4"
format: html
editor: visual
---

## Problem 1

```{r}
simulate_pi <- function(n = 5e6) {
  inside <- 0
  for (i in 1:n) {
    x <- runif(1, -1, 1)
    y <- runif(1, -1, 1)
    if (x^2 + y^2 <= 1) inside <- inside + 1
  }
  4 * inside / n
}

set.seed(1)
simulate_pi(5e6)
```

Run the profvis

```{r}
library(profvis)

profvis({
  simulate_pi()
})


```

The flame graph shows that `runif()` dominates the execution time. The entire `simulate_pi()` function took 4280 ms, while `runif()`accounted for 4020 ms. Therefore, approximately 4020/4280=93.9% of the total runtime was spent in `runif()`, leaving about 6.1% for everything else in the loop. This occurs because `runif(1, -1, 1)` is called twice per iteration, resulting in 10 million calls when n=5×10\^6.

```{r}
simulate_pi_vec <- function(n = 5e6) {
  x <- runif(n, -1, 1)
  y <- runif(n, -1, 1)
  
  inside <- x^2 + y^2 <= 1
  
  4 * mean(inside)
}

bench::mark(
  loop = simulate_pi(),
  vectorized = simulate_pi_vec(),
  relative = TRUE,
  check = FALSE
)
```

Speed up from 4.82s to 617.64ms

## Problem 2

```{r}
make_kernel <- function(coords, bandwidth) {
  d <- as.matrix(dist(coords))
  K <- exp(-(d / bandwidth)^2)
  K / rowSums(K)
}

# Advance p0 forward nsteps times: p(s+1) = p(s) %*% K.
simulate_spread <- function(coords, p0, nsteps = 300, bandwidth = 0.15) {
  p <- p0
  for (s in seq_len(nsteps)) {
    K <- make_kernel(coords, bandwidth)
    p <- as.vector(p %*% K)
    p <- p / sum(p)
  }
  p
}

set.seed(2)
k <- 1000
coords <- cbind(runif(k), runif(k))
p0 <- rep(1 / k, k)

p_final <- simulate_spread(coords, p0)
summary(p_final)

```

use profvis

```{r}

profvis({
  result_old <- simulate_spread(coords, p0)
})


# Run the original version and time it
time_old <- system.time({
  result_old <- simulate_spread(coords, p0)
})

time_old

```

A large amount of time is in make_kernel(coords, bandwidth), especially in the as.matrix part, because The reason is that the same 1000 × 1000 matrix is rebuilt 300 times.

fix it

```{r}
simulate_spread_fixed <- function(coords, p0,
                                  nsteps = 300,
                                  bandwidth = 0.15) {
  
  # K does not depend on s, so calculate it only once
  K <- make_kernel(coords, bandwidth)
  
  p <- p0
  
  for (s in seq_len(nsteps)) {
    p <- as.vector(p %*% K)
    p <- p / sum(p)
  }
  
  p
}
```

```{r}
# Original
result_old <- simulate_spread(coords, p0)

# Fixed
result_new <- simulate_spread_fixed(coords, p0)

# Confirm that the results are the same
all.equal(result_old, result_new)
```

The result is unchanged. Now benchmark:

```{r}
library(microbenchmark)

microbenchmark(
  original = simulate_spread(coords, p0),
  fixed = simulate_spread_fixed(coords, p0),
  times = 5
)
```

The original function had a median run time of about 6.45 seconds, while the optimized function took about 0.45 seconds. The optimized version is approximately 14.3 times faster. This suggests that about 6.0 seconds

## Problem 3

```{r}
fit_poisson_nr <- function(X, y, maxit = 50, tol = 1e-8) {
  beta <- rep(0, ncol(X))
  beta[1] <- log(mean(y))

  for (it in 1:maxit) {
    mu <- exp(drop(X %*% beta))
    score <- crossprod(X, y - mu)
    info <- crossprod(X * mu, X)
    step <- solve(info, score)

    beta <- beta - drop(step)

    if (max(abs(step)) < tol) break
  }

  list(coefficients = beta, iterations = it)
}
```

```{r}
source("/Users/zhoumo/Downloads/poisson-nr.R")

fit_poisson_nr(X, y)
```

```{r}
traceback()
```

source it again after change

```{r}
source("/Users/zhoumo/Downloads/poisson-nr.R")

fit_poisson_nr(X, y)
```

I found the beta \<- beta - drop(step) line is problematic. The log likelihood is decreasing each time. We should change into +.

## Problem 4

```{r}
source("/Users/zhoumo/Downloads/predict-scaling.R")
fit <- fit_poisson(X, y)
predict_counts(fit, X)
predict_counts(fit, X[1, , drop = FALSE])
```

The problem line is Z \<- scale(newdata)

```{r}
fit_poisson <- function(X, y, maxit = 50, tol = 1e-8) {
  Z <- scale(X)
  
  # Save training-data scaling parameters
  center <- attr(Z, "scaled:center")
  scale_value <- attr(Z, "scaled:scale")
  
  Zd <- cbind(1, Z)
  
  beta <- rep(0, ncol(Zd))
  beta[1] <- log(mean(y))
  
  for (it in 1:maxit) {
    mu <- exp(drop(Zd %*% beta))
    score <- crossprod(Zd, y - mu)
    info <- crossprod(Zd * mu, Zd)
    step <- solve(info, score)
    
    beta <- beta + drop(step)
    
    if (max(abs(step)) < tol) break
  }
  
  list(
    coefficients = beta,
    iterations = it,
    center = center,
    scale = scale_value
  )
}


predict_counts <- function(fit, newdata) {
  # Use the training-data center and scale
  Z <- scale(
    newdata,
    center = fit$center,
    scale = fit$scale
  )
  
  Zd <- cbind(1, Z)
  
  exp(drop(Zd %*% fit$coefficients))
}


set.seed(1)

n <- 400

X <- cbind(
  x1 = rnorm(n, 50, 10),
  x2 = rnorm(n, 2, 0.5)
)

beta_true <- c(-1, 0.9, 0.3)

Z_true <- scale(X)

y <- rpois(
  n,
  exp(drop(cbind(1, Z_true) %*% beta_true))
)


# Fit
fit <- fit_poisson(X, y)

# New clinic
new_clinic <- X[1, , drop = FALSE]

# Prediction
prediction <- predict_counts(fit, new_clinic)

prediction

dat <- data.frame(
  y = y,
  x1 = X[, 1],
  x2 = X[, 2]
)
# Compare with glm()
glm_fit <- glm(
  y ~ x1 + x2,
  data=dat,
  family = poisson()
)

glm_prediction <- predict(
  glm_fit,
  newdata = data.frame(
    x1 = X[1, 1],
    x2 = X[1, 2]
  ),
  type = "response"
)

prediction
glm_prediction


# Check whether they agree
all.equal(
  as.numeric(prediction),
  as.numeric(glm_prediction),
  tolerance = 1e-6
)
```
